---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# moat

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

MOAT is the Microbiome Omics Audit Toolkit. Before crossing into microbiome
analysis, check the moat.

`moat` audits microbiome and omics study designs before downstream statistical
or machine-learning analysis. It checks whether outcome, batch, covariates,
repeated measures, correction identifiability, and microbiome distance
structure create risks that should be reported before differential abundance,
PERMANOVA, ordination, or predictive modeling.

The package is a study-audit layer. It does not silently correct counts, remove
samples, or choose the final model. Instead, it returns structured diagnostics,
risk levels, warnings, plots, and analysis-plan suggestions that analysts can
review and report.

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("xec-cm/moat")
```

## Quick Start

```{r quick-start}
library(moat)

data("toy_moat")

audit <- moat(
    toy_moat,
    outcome = "outcome",
    batch = "batch",
    distances = "bray",
    n_perm = 99,
    verbose = FALSE
)

summary(audit)
```

The audit object keeps module-level diagnostics:

```{r inspect-audit}
audit$design[, c("variable", "role", "effect_size_name", "effect_size", "risk")]
audit$batch$summary[, c("distance", "outcome_r2", "batch_r2", "risk")]
audit$correction$feasibility
audit$leakage$recommended_cv
```

`plan_analysis()` translates the audit into downstream analysis guidance:

```{r analysis-plan}
plan_analysis(audit)
```

## Plotting

`plot_design()` visualizes the outcome distribution across an audited
categorical variable. Empty cells are highlighted because they often signal
confounding or complete separation.

`plot_variance()` visualizes PERMANOVA R2 by term, making batch dominance easy
to spot.

```{r plotting, eval = FALSE}
plot_design(audit, variable = "batch")
plot_variance(audit, distance = "bray")
```

## What MOAT Checks

- **Design confounding:** whether outcome is associated with batch or
  covariates in metadata.
- **Correction feasibility:** whether batch adjustment is statistically
  identifiable from the observed design.
- **Microbiome batch effects:** whether batch explains microbiome variation
  using PERMANOVA, dispersion, and ordination diagnostics.
- **Validation leakage:** whether repeated measures, batch-outcome association,
  or time structure require grouped or time-aware validation.
- **Analysis planning:** recommended formulas, permutation restrictions,
  validation schemes, batch strategy, and sensitivity analyses.

## Scientific Position

`moat` is intentionally conservative. A high-risk audit does not prove that
an analysis is invalid, and a low-risk audit does not prove that all downstream
models are safe. The package is designed to make design limitations explicit so
that conclusions are interpreted with the right sensitivity analyses and
reporting context.

## Development Checks

```{r development, eval = FALSE}
testthat::test_local()
devtools::check()
BiocCheck::BiocCheck()
```
