---
title: "Why pre-analysis audit changes interpretation"
author:
  - name: Francesc Catala-Moll
    affiliation: IrsiCaixa AIDS Research Institute
package: moat
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Why pre-analysis audit changes interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.2
)
```

# Why Interpretation Needs An Audit

A microbiome result can look biologically plausible while still being driven by
study design, batch structure, or validation leakage. MOAT is meant to make
those risks visible before the downstream model becomes the story.

This vignette shows three compact examples where a naive interpretation changes
after checking the audit:

1. a beta-diversity signal dominated by batch,
2. outcome information already present in metadata,
3. validation leakage from repeated subjects or batch-outcome alignment.

```{r load-package}
library(moat)
data("toy_moat")
```

# Batch-Dominated Beta Diversity

A common first pass is to ask whether microbiome composition is associated with
the outcome. The code below fits an outcome-only PERMANOVA on Bray-Curtis
distances.

```{r naive-permanova}
metadata <- as.data.frame(SummarizedExperiment::colData(toy_moat))
distance <- compute_biome_distance(toy_moat, distance = "bray")

naive <- check_permanova(
  distance,
  metadata = metadata,
  outcome = "outcome",
  n_perm = 99
)

naive$terms[, c("term", "r2", "p_value")]
```

Read alone, this can encourage an outcome-centered interpretation. The audit
adds the batch variable and reports the outcome and batch contributions
together.

```{r batch-audit}
audit <- moat(
  toy_moat,
  outcome = "outcome",
  batch = "batch",
  distances = "bray",
  n_perm = 99,
  verbose = FALSE
)

audit$batch$summary[, c(
  "distance",
  "outcome_r2",
  "batch_r2",
  "batch_dominance_score",
  "permanova_risk",
  "order_sensitivity_risk",
  "risk"
)]
```

When batch explains as much or more variation than the outcome, the result
should be reported as a batch-sensitive finding rather than a clean biological
separation. The practical next step is to carry the audit into the analysis
plan.

```{r batch-plan}
plan <- plan_analysis(audit)

plan$permanova
plan$batch_strategy
```

# Metadata-Only Outcome Predictability

Microbiome models can also inherit information from metadata. In this small
example, center is strongly aligned with outcome before any microbiome feature
is used.

```{r metadata-risk-data}
metadata_confounded <- data.frame(
  outcome = rep(c("Control", "Disease"), each = 20),
  center = rep(c("Center_A", "Center_B"), each = 20),
  age_group = rep(c("young", "old"), times = 20)
)

table(metadata_confounded$center, metadata_confounded$outcome)
```

`check_design()` stores a metadata-only predictability diagnostic as an
attribute. A high value means downstream microbiome signatures may partly
reflect design or validation structure.

```{r metadata-predictability}
design <- check_design(
  metadata_confounded,
  outcome = "outcome",
  batch = "center",
  covariates = "age_group"
)

design[, c("variable", "role", "effect_size_name", "effect_size", "risk")]

metadata_predictability <- attr(design, "metadata_predictability")
metadata_predictability[c("status", "risk", "cv_balanced_accuracy", "recommendations")]
```

The interpretation changes from "metadata are harmless annotations" to
"metadata already carries outcome information". Any microbiome model should
report this risk and use validation that does not let center or batch structure
stand in for biology.

# Validation Leakage

Leakage appears when the validation split lets the model reuse structure that
will not generalize. Repeated subjects are a simple example: if samples from the
same subject appear in both train and test folds, row-wise cross-validation can
look better than subject-level prediction.

```{r repeated-subjects}
repeated_biome <- toy_moat
SummarizedExperiment::colData(repeated_biome)$subject <- rep(
  paste0("S", seq_len(ncol(repeated_biome) / 2)),
  each = 2
)
SummarizedExperiment::colData(repeated_biome)$timepoint <- rep(
  c("baseline", "followup"),
  times = ncol(repeated_biome) / 2
)

repeated_metadata <- as.data.frame(SummarizedExperiment::colData(repeated_biome))
head(repeated_metadata)
```

`check_leakage()` turns that metadata structure into validation guidance.

```{r leakage-check}
leakage <- check_leakage(
  repeated_metadata,
  outcome = "outcome",
  subject = "subject",
  batch = "batch",
  time = "timepoint"
)

leakage[c("risk", "recommended_cv", "recommendations")]
```

The same recommendation is carried by the full audit and the analysis plan.

```{r leakage-plan}
leakage_audit <- moat(
  repeated_biome,
  outcome = "outcome",
  batch = "batch",
  subject = "subject",
  time = "timepoint",
  distances = "bray",
  n_perm = 99,
  verbose = FALSE
)

leakage_plan <- plan_analysis(leakage_audit)

leakage_plan$ml_validation
leakage_plan$permutation
```

# Practical Reporting Pattern

For each downstream analysis, report the naive target next to the audit-guided
qualification:

- report outcome R2 together with batch R2 before interpreting beta diversity,
- report metadata-only predictability before claiming microbiome prediction,
- report leakage-aware validation when repeated subjects, time, or batch
  structure are present,
- use `plan_analysis()` to turn the audit into formulas, validation schemes,
  and sensitivity checks.

# Session Information

```{r session-info}
sessionInfo()
```
