---
title: "Validation leakage in microbiome machine learning"
author:
  - name: Francesc Catala-Moll
    affiliation: IrsiCaixa AIDS Research Institute
package: moat
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Validation leakage in microbiome machine learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.2
)
```

# Why Leakage Matters

Microbiome machine-learning models often work with samples, but the scientific
unit of generalization may be a subject, hospital, sequencing run, visit, or
study site. If train and test folds split those units incorrectly, the model can
learn repeated-subject structure or batch signatures instead of biology.

This vignette focuses on three common leakage patterns:

1. repeated measurements from the same subject,
2. outcome labels aligned with batch or center,
3. time-ordered samples from the same subjects.

MOAT does not train the machine-learning model. It audits metadata before
modeling and recommends validation schemes that make leakage harder to miss.

```{r load-package}
library(moat)
data("toy_moat")
```

# Row-Wise Cross-Validation Can Be Overoptimistic

Row-wise cross-validation treats each sample as independent. That assumption is
fragile when several rows come from the same person. In the small simulated
dataset below, every subject contributes two samples.

```{r repeated-data}
repeated_biome <- toy_moat
SummarizedExperiment::colData(repeated_biome)$subject <- rep(
  paste0("S", seq_len(ncol(repeated_biome) / 2)),
  each = 2
)
SummarizedExperiment::colData(repeated_biome)$timepoint <- rep(
  c("baseline", "followup"),
  times = ncol(repeated_biome) / 2
)
metadata <- as.data.frame(SummarizedExperiment::colData(repeated_biome))

head(metadata)
```

If ordinary row-wise folds are used, one sample from a subject can be in the
training fold while another sample from the same subject is in the test fold.
The test fold is then no longer an independent subject-level evaluation.

```{r repeated-check}
repeated <- check_repeated_measures(metadata, subject = "subject")

repeated[c(
  "n_samples",
  "n_subjects",
  "n_repeated_subjects",
  "max_samples_per_subject",
  "risk",
  "recommended_cv"
)]
```

The actionable recommendation is to keep samples from the same subject together
inside each resampling split.

```{r repeated-subject-counts}
head(repeated$samples_per_subject)
```

# Use Grouped Cross-Validation

`check_leakage()` combines repeated-measure, batch, and temporal diagnostics.
For repeated measures without an explicit time variable, the recommended scheme
is grouped cross-validation by subject.

```{r grouped-cv}
leakage_grouped <- check_leakage(
  metadata,
  outcome = "outcome",
  subject = "subject"
)

leakage_grouped$recommended_cv
leakage_grouped$recommendations
```

In downstream modeling, that means fold assignment should happen at subject
level first. All rows from a subject should move together into train or test.

# Leave-One-Center-Out For Batch Leakage

Another failure mode appears when a center, site, or sequencing batch is aligned
with the outcome. A model can then perform well by recognizing center-specific
microbiome or technical signatures.

```{r confounded-data}
confounded_metadata <- data.frame(
  outcome = rep(c("Control", "Disease"), each = 10),
  batch = rep(c("Center_A", "Center_B"), each = 10)
)

table(confounded_metadata$batch, confounded_metadata$outcome)
```

Here each batch level contains only one outcome group. A random row-wise split
can put both centers into both train and test, making center recognition look
like outcome prediction.

```{r batch-leakage}
batch_leakage <- check_leakage(
  confounded_metadata,
  outcome = "outcome",
  batch = "batch"
)

batch_leakage$risk
batch_leakage$recommended_cv
batch_leakage$batch_leakage$summary
```

For this scenario, a leave-one-batch-out strategy is a useful stress test. It
asks whether the model still works when a whole center or batch is held out.

# Time-Aware Validation

Longitudinal data add another constraint: train folds should not silently use
future samples from the same subject to predict earlier or held-out samples.
When repeated subjects span timepoints, MOAT recommends a grouped
time-aware strategy.

```{r temporal-leakage}
leakage_time <- check_leakage(
  metadata,
  outcome = "outcome",
  subject = "subject",
  time = "timepoint"
)

leakage_time$risk
leakage_time$recommended_cv
leakage_time$temporal_leakage[c(
  "n_timepoints",
  "subjects_with_multiple_timepoints",
  "risk"
)]
```

This does not replace a full temporal validation design. It flags that ordinary
row-wise cross-validation is not enough for the metadata structure supplied.

# Full Audit Integration

The same leakage recommendations flow through `moat()` and
`plan_analysis()`. This lets the validation choice sit next to design, batch,
and correction diagnostics instead of being handled as an afterthought.

```{r full-audit}
audit <- moat(
  repeated_biome,
  outcome = "outcome",
  subject = "subject",
  time = "timepoint",
  distances = "bray",
  n_perm = 99,
  verbose = FALSE
)

summary(audit)
```

```{r plan-validation}
plan <- plan_analysis(audit)

plan$ml_validation
plan$permutation
```

# Preprocessing Leakage Checklist

Validation leakage is not only about fold assignment. Any operation that learns
from the full dataset before splitting can leak information into test folds.

```{r preprocessing-checklist}
audit$leakage$preprocessing_checklist
```

Before fitting a model, check that the following steps are estimated inside
each training fold and then applied to the held-out fold:

- feature filtering,
- prevalence filtering,
- normalization or scaling parameters,
- imputation,
- feature selection,
- hyperparameter tuning.

# Practical Checklist For ML Users

Use this checklist before interpreting microbiome ML performance:

- Are multiple samples from the same subject kept in the same fold?
- If outcome is aligned with center or batch, is leave-one-center-out or
  leave-one-batch-out performance reported?
- For longitudinal data, does validation respect subject grouping and time?
- Are preprocessing, filtering, scaling, and feature selection nested inside
  resampling?
- Are ordinary CV and leakage-aware CV results shown side by side when risk is
  moderate or high?

# Session Information

```{r session-info}
sessionInfo()
```
