---
title: "Auditing microbiome study designs with MOAT"
author:
  - name: Francesc Catala-Moll
    affiliation: IrsiCaixa AIDS Research Institute
package: moat
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Auditing microbiome study designs with MOAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)
```

# Why Audit Microbiome Studies?

Microbiome analyses are sensitive to study design. Outcome groups can be
confounded with center, sequencing run, extraction kit, visit, or subject
identity. When this happens, downstream analyses can report biological
differences that are partly or entirely driven by technical or design
structure.

MOAT audits those risks before analysis. It does not automatically correct
counts or decide the final model. Instead, it returns diagnostics, warnings,
plots, and analysis-plan recommendations that help make the study design
explicit. Before crossing into microbiome analysis, check the moat.

# Load the Toy Dataset

```{r load-package}
library(moat)
data("toy_moat")

toy_moat
```

The example dataset is a small simulated `SummarizedExperiment` with count data
and sample metadata.

```{r inspect-metadata}
head(SummarizedExperiment::colData(toy_moat))
```

# Run the Audit

For examples and vignettes we use a small number of permutations. Real analyses
should generally use a larger value, such as the default `999`.

```{r run-audit}
audit <- moat(
  toy_moat,
  outcome = "outcome",
  batch = "batch",
  distances = "bray",
  n_perm = 99,
  verbose = FALSE
)

summary(audit)
```

For compact reporting, `module_risks()` extracts the module-level status and
main reason without requiring users to inspect the nested audit object.

```{r module-risks}
module_risks(audit)
```

# Interpret Design Risk

The design module checks whether metadata variables are associated with the
outcome. For categorical variables it stores a contingency table, Cramer's V,
empty-cell counts, and a conservative risk label.

```{r design-table}
audit$design[, c(
  "variable",
  "role",
  "variable_type",
  "effect_size_name",
  "effect_size",
  "empty_cells",
  "risk"
)]
```

`plot_design()` shows the outcome distribution across an audited categorical
variable. Empty cells are outlined because they can indicate confounding or
complete separation.

```{r plot-design}
plot_design(audit, variable = "batch")
```

# Interpret Correction Feasibility

Batch adjustment is only meaningful when outcome and batch are not perfectly
aliased. The correction module summarizes whether ordinary adjustment appears
identifiable from the metadata.

```{r correction-feasibility}
audit$correction$feasibility
audit$correction$recommendations
```

This diagnostic is not a command to correct the data. It is a warning about
what kinds of downstream adjustment are statistically defensible.

Thresholds used by all risk modules are documented in the companion article
`vignette("risk-thresholds", package = "moat")` and exposed as a table by
`risk_thresholds()`.

# Interpret Batch Risk

The batch module combines distance-based PERMANOVA, dispersion diagnostics, and
ordination-axis association checks. The summary table reports how much
microbiome variation is attributed to the outcome and to batch.

```{r batch-summary}
audit$batch$summary[, c(
  "distance",
  "outcome_r2",
  "batch_r2",
  "batch_dominance_score",
  "permanova_risk",
  "permdisp_risk",
  "pcoa_risk",
  "risk"
)]
```

`plot_variance()` visualizes the PERMANOVA R2 terms. A batch term that exceeds
the outcome term is a strong signal that the downstream analysis needs explicit
batch sensitivity checks.

```{r plot-variance}
plot_variance(audit, distance = "bray")
```

# Interpret Leakage Risk

Validation leakage can occur when samples from the same subject, batch, or time
structure are split across train and test folds in a way that makes prediction
look better than it is. MOAT reports the recommended cross-validation
scheme based on the metadata supplied to `moat()`.

```{r leakage-summary}
audit$leakage$recommended_cv
audit$leakage$recommendations
```

# Generate an Analysis Plan

`plan_analysis()` translates the audit into recommended formulas, validation
schemes, batch strategy, and sensitivity analyses.

```{r analysis-plan}
plan_analysis(audit)
```

The plan should be read as a structured starting point. Analysts still need to
choose methods appropriate for their scientific question, data distribution,
and experimental design.

# Session Information

```{r session-info}
sessionInfo()
```
