---
title: "How MOAT risk levels are assigned"
author:
  - name: Francesc Catala-Moll
    affiliation: IrsiCaixa AIDS Research Institute
package: moat
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{How MOAT risk levels are assigned}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Conservative Screening Heuristics

MOAT assigns qualitative risk levels before downstream microbiome analyses are
run. These levels are conservative screening heuristics: they are designed to
make design limitations visible, not to prove that an analysis is valid or
invalid.

The common risk levels are:

- `low`: no configured warning threshold was met.
- `moderate`: a visible design or diagnostic signal should be inspected and
  reported.
- `high`: a strong design or diagnostic signal requires guarded interpretation
  or sensitivity analyses.
- `critical`: the observed design can make naive adjustment or interpretation
  non-identifiable.
- `unknown`: the module was skipped, unavailable, or did not have enough
  information to score.

The programmatic reference is available from `risk_thresholds()`.

```{r thresholds}
library(moat)

risk_thresholds()
```

# Design Risk

Design risk checks whether outcome groups are associated with batch variables
or covariates in metadata. Categorical metadata uses contingency-table
diagnostics such as Cramer's V, empty cells, minimum cell count, and complete
separation. Continuous metadata uses standardized mean differences, p-values,
and outcome imbalance.

# Batch-Space Risk

Batch-space risk is evaluated on microbiome distances with PERMANOVA,
dispersion diagnostics, and PCoA-axis associations. A high batch-space score can
come from large batch R2, batch dominance over outcome R2, low PERMANOVA or
PERMDISP p-values, or strong alignment between batch variables and ordination
axes.

# Correction Identifiability

Correction feasibility is based on batch-by-outcome positivity and adjustment
model diagnostics. Complete separation, rank deficiency, and severe
collinearity are treated conservatively because ordinary batch adjustment may
remove biology or create artifacts when the design is not identifiable.

# Metadata-Only Predictability

Metadata-only outcome predictability is scored with balanced accuracy from
cross-validation when possible, falling back to apparent balanced accuracy when
cross-validation cannot be estimated. Strong metadata-only prediction indicates
that downstream microbiome models may be vulnerable to design confounding or
validation leakage.

# Validation Leakage Risk

Leakage risk covers repeated subjects, batch-outcome association, and temporal
structure. Repeated measures and timepoints can make ordinary cross-validation
optimistic, so MOAT recommends grouped or time-aware validation when these risks
are detected.

# Global Risk Aggregation

The overall risk is the maximum normalized risk across design, batch,
correction, and leakage modules. Overall reasons are drawn from the modules that
reached the maximum risk level, while recommendations are collected from module
outputs and de-duplicated.
