---
title: "safebiome"
author:
  - name: Francesc Catala-Moll
    affiliation: IrsiCaixa AIDS Research Institute
package: safebiome
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{safebiome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

`safebiome` is an R package for auditing microbiome study designs before
downstream statistical analysis.

The package is in early development. Its planned scope includes checks for
design confounding, batch effects, correction feasibility, and validation
leakage.

# Installation check

```{r load-package}
library(safebiome)
```

# Planned workflow

The intended user workflow is:

```{r planned-workflow, eval=FALSE}
audit <- check_biome(
    x,
    outcome = "condition",
    batch = c("center", "sequencing_run"),
    covariates = c("age", "sex", "antibiotic_use"),
    subject = "patient_id"
)

summary(audit)
plan_analysis(audit)
```

These user-facing functions will be implemented in upcoming development
milestones.

# Session information

```{r session-info}
sessionInfo()
```
